---
title: "PHDS Data Visualization Workshop: Data Carpentry Dataset"
author: "Lukas M. Weber"
date: "2024-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This session is adapted from the Data Carpentry lesson "Data visualization with ggplot2": https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html

The examples in this session show how to generate and customize plots using ggplot2 in R, using an ecology dataset containing survey data on animal species diversity and weights within a study site.


# Example

## Load data

We load the dataset in .csv format from GitHub. The saved .csv file on GitHub includes all preprocessing from the earlier chapters in the Data Carpentry materials.

Code for the preprocessing steps is also included in the [previous script](https://github.com/lmweber/PHDS-data-visualization-2024/blob/main/code/PHDS_Part_2a_Data_Carpentry_Preprocessing.Rmd), available on GitHub.

```{r, message=FALSE}
library(tidyverse)
library(here)
library(ggplot2)
library(hexbin)
```

```{r}
# download saved dataset from GitHub
download.file(
  url = "https://raw.githubusercontent.com/lmweber/PHDS-data-visualization-2024/main/data/surveys_complete.csv", 
  destfile = here("data/surveys_complete.csv")
)
```

```{r}
# load data
surveys_complete <- read_csv(here("data/surveys_complete.csv"))
```

Have a look at the dataset:

```{r}
head(surveys_complete)
```


## Scatter plot

Generate an initial scatter plot. Note the first two arguments to the `ggplot2()` function (`data` and `mapping = aes()`), the syntax for `aes()`, the use of `geom_point()`, and the `+` syntax for combining elements of the plot:

```{r}
# generate an initial scatter plot
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) + 
  geom_point()
```


### Alternative `geom`:

We can also specify a different `geom`:

```{r}
# 'geom_hex()'
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) + 
  geom_hex()
```


### Modifying the plot

We can modify the plot by providing additional arguments:

The `alpha` argument specifies a level of transparency:

```{r}
# 'alpha' argument
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) + 
  geom_point(alpha = 0.1)
```

The `color` argument sets colors:

```{r}
# 'color' argument
ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) + 
  geom_point(alpha = 0.1, color = "blue")
```

We can also use the `color` argument within `aes()` to color points by the values of a variable. Note that we can specify `color` within `aes()` either at the top level (within the `ggplot2()` call) or within the `geom_point()` call. This gives flexibility for setting colors only for certain `geoms` or for the whole plot.

```{r}
# set 'color' using categorical variable

ggplot(surveys_complete, aes(x = weight, y = hindfoot_length)) + 
  geom_point(alpha = 0.1, aes(color = species_id))

ggplot(surveys_complete, aes(x = weight, y = hindfoot_length, 
                             color = species_id)) + 
  geom_point(alpha = 0.1)
```


## Boxplots

Generate boxplots:

```{r}
# generate boxplots
ggplot(surveys_complete, aes(x = species_id, y = weight)) + 
  geom_boxplot()
```

Add points and additional formatting arguments to the boxplots:

```{r}
# boxplots with points and formatting
ggplot(surveys_complete, aes(x = species_id, y = weight)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(alpha = 0.3, color = "tomato")
```


## Violin plots

Alternatively, we can use violin plots:

```{r}
# generate violin plots
ggplot(surveys_complete, aes(x = species_id, y = weight)) + 
  geom_violin()
```


## Line plots

To demonstrate line plots for time series data, we calculate an additional variable:

```{r}
# calculate number of counts per year for each genus
yearly_counts <- 
  surveys_complete |> 
  count(year, genus)

head(yearly_counts)
```

To generate line plots, we can specify `group` in `aes()`. However, this does not allow us to identify the lines.

```{r}
# generate line plot
ggplot(yearly_counts, aes(x = year, y = n, group = genus)) + 
  geom_line()
```

We can use `color` to identify the lines. Note that setting `color` in `aes()` also automatically groups the data.

```{r}
# color by genus variable
ggplot(data = yearly_counts, aes(x = year, y = n, color = genus)) + 
  geom_line()
```


## Facetting

In ggplot, `facetting` refers to splitting plots into multiple panels according to some variable. This can be very useful, and requires a special syntax.

Here, we use `facet_wrap()` to facet the line plots by genus:

```{r}
# facet by genus
ggplot(yearly_counts, aes(x = year, y = n)) + 
  geom_line() + 
  facet_wrap(vars(genus))
```

We can also include color. However, in this case we have some redundancy between the facetting and the colors.

```{r}
# facet by genus and include color
ggplot(yearly_counts, aes(x = year, y = n, color = genus)) + 
  geom_line() + 
  facet_wrap(vars(genus))
```

Alternatively, we can use facetting and further split the lines by sex. To do this, we need to do some further data manipulation.

```{r}
# split counts by sex
yearly_sex_counts <- 
  surveys_complete |> 
  count(year, genus, sex)
```

Now we can create a facetted plot with lines split and colored by the `sex` variable:

```{r}
# facetted plot
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_wrap(vars(genus))
```

We can also facet by multiple variables using `facet_grid()`. In this case, the plot may be getting quite busy and more difficult to read.

```{r}
# facet by multiple variables using 'facet_grid()'
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_grid(rows = vars(sex), cols = vars(genus))
```

Alternatively, facets may be organized by either rows or columns:

```{r}
# facet by rows with one column
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_grid(rows = vars(genus))
```

```{r}
# facet by columns with one row
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_grid(cols = vars(genus))
```


## Themes

Themes provide a convenient way to format plots in ggplot2.

For example, `theme_bw()` adjusts the formatting to show white backgrounds, black borders, gray gridlines, and additional default settings. There are also several other themes available.

```{r}
# demonstrate 'theme_bw()'
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_wrap(vars(genus)) + 
  theme_bw()
```


## Additional customizations

Additional functions are available to further customize the plot. For example, we can add more informative plot titles and axis titles.

```{r}
# specify plot title and axis titles
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_wrap(vars(genus)) + 
  labs(title = "Observed genera through time", 
        x = "Year of observation", 
       y = "Number of individuals") + 
  theme_bw()
```

The `theme()` function can be used to add further customizations, such as font size. Note the special syntax using `element_text()`.

```{r}
# using 'theme()' to adjust font size
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_wrap(vars(genus)) + 
  labs(title = "Observed genera through time", 
       x = "Year of observation", 
       y = "Number of individuals") + 
  theme_bw() + 
  theme(text = element_text(size = 16))
```

Numerous additional options are available for detailed customizations:

```{r}
# demonstrate additional options for 'theme()'
ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_wrap(vars(genus)) + 
  labs(title = "Observed genera through time", 
       x = "Year of observation", 
       y = "Number of individuals") + 
  theme_bw() + 
  theme(axis.text.x = element_text(color = "grey20", size = 12, angle = 90, 
                                   hjust = 0.5, vjust = 0.5), 
        axis.text.y = element_text(color = "gray20", size = 12), 
        strip.text = element_text(face = "italic"), 
        text = element_text(size = 16))
```


## Using 'patchwork' for multiple panels

Here, we demonstrate the use of the `patchwork` package to arrange multiple plots in panels.

First, we load the `patchwork` package:

```{r, message=FALSE}
library(patchwork)
```

The `patchwork` package uses a specific syntax (`|` for columns, `/` for rows):

```{r}
# create plots and assign the plots to variables

plot_weight <- ggplot(surveys_complete, aes(x = species_id, y = weight)) + 
  geom_boxplot() + 
  labs(x = "Species", y = expression(log[10](Weight))) + 
  scale_y_log10()

plot_count <- ggplot(yearly_counts, aes(x = year, y = n, color = genus)) + 
  geom_line() + 
  labs(x = "Year", 
       y = "Abundance")
```

```{r}
# display plots in rows
plot_weight / plot_count + 
  plot_layout(heights = c(3, 2))
```

```{r}
# display plots in columns
plot_weight | plot_count
```


## Export plots

Finally, we can export or save plots using the `ggsave()` function. Usually we will save plots in either `.png` or `.pdf` format. The `ggsave()` function detects the format automatically from the filename.

```{r}
# generate plot
my_plot <- ggplot(yearly_sex_counts, aes(x = year, y = n, color = sex)) + 
  geom_line() + 
  facet_wrap(vars(genus)) + 
  labs(title = "Observed genera through time", 
       x = "Year of observation", 
       y = "Number of individuals") + 
  theme_bw() +
  theme(axis.text.x = element_text(color = "gray20", size = 12, angle = 90, 
                                   hjust = 0.5, vjust = 0.5), 
        axis.text.y = element_text(color = "gray20", size = 12), 
        text = element_text(size = 16))
```

```{r}
# display plot
my_plot
```

Save plot in .png format:

```{r}
# save plot in .png format
ggsave(here("plots/my_plot.png"), my_plot, width = 15, height = 10)
```

Save plot in .pdf format:

```{r}
# save plot in .png format
ggsave(here("plots/my_plot.pdf"), my_plot, width = 15, height = 10)
```

We can also specify the resolution for .png format:

```{r}
# save combined plot with resolution argument in .png format
plot_combined <- 
  plot_weight / plot_count + 
  plot_layout(heights = c(3, 2))

ggsave(here("plots/plot_combined.png"), plot_combined, width = 10, dpi = 300)
```
